---
title: "Merging Format 5 Binary Dosage Files"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Merging Format 5 Binary Dosage Files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE}
library(BinaryDosage)
```

# Introduction

The `mergebd` function merges two or more Format 5 binary dosage files into a
single Format 5 output file. The merge type is detected automatically from the
input files.

- **Subject merge** — subject IDs do not overlap across files. The output
  contains all subjects from every input file and the SNPs common to all files.
- **SNP merge** — SNP IDs do not overlap across files. The output contains all
  SNPs from every input file and the subjects common to all files.

If both subject IDs and SNP IDs overlap across files an error is returned,
since the merge type is ambiguous.

SNPs are identified by chromosome, position, reference allele, and alternate
allele, regardless of the SNP ID format stored in each file.

The function takes the following parameters.

- `bdose_files` — character vector of paths to the input `.bdose` files (at
  least two). The companion `.bdi` file for each is expected at
  `paste0(bdose_files[i], ".bdi")`.
- `bdose_file` — path for the output `.bdose` file. The companion `.bdi` file
  is written automatically to `paste0(bdose_file, ".bdi")`.

# Setup

The examples below use the bgzipped VCF file included with the package,
*set1a.vcf.gz*, which contains data for 60 subjects and 10 SNPs on chromosome
1. All output files are written to a temporary directory.

```{r setup_files, message = FALSE, warning = FALSE}
vcf_file   <- system.file("extdata", "set1a.vcf.gz", package = "BinaryDosage")
bdose_full <- file.path(tempdir(), "full.bdose")

vcftobd(vcffile = vcf_file, bdose_file = bdose_full)
bd_full <- getbdinfo(bdose_full)

cat("Subjects:", nrow(bd_full$samples), "\n")
cat("SNPs:    ", nrow(bd_full$snps),    "\n")
```

# Subject merge

A subject merge combines files that cover different subjects but the same (or
overlapping) set of SNPs. The output contains all subjects and the SNPs common
to every input file.

The example splits the 60-subject file into two 30-subject files using
`subsetbd`, then merges them back together.

```{r subject_merge, message = FALSE, warning = FALSE}
bdose_a   <- file.path(tempdir(), "set_a.bdose")
bdose_b   <- file.path(tempdir(), "set_b.bdose")
bdose_out <- file.path(tempdir(), "merged_subjects.bdose")

sids <- bd_full$samples$sid

subsetbd(bdfiles    = bdose_full,
         bdose_file = bdose_a,
         subjectids = sids[1:30])

subsetbd(bdfiles    = bdose_full,
         bdose_file = bdose_b,
         subjectids = sids[31:60])

mergebd(bdose_files = c(bdose_a, bdose_b),
        bdose_file  = bdose_out)

bd_a   <- getbdinfo(bdose_a)
bd_b   <- getbdinfo(bdose_b)
bd_out <- getbdinfo(bdose_out)

cat("File A subjects:", nrow(bd_a$samples),   "\n")
cat("File B subjects:", nrow(bd_b$samples),   "\n")
cat("Merged subjects:", nrow(bd_out$samples), "\n")
cat("Merged SNPs:    ", nrow(bd_out$snps),    "\n")
```

The merged file contains all 60 subjects and all 10 SNPs.

## Verifying subject order

The subjects in the merged file appear in input-file order: all subjects from
the first file followed by all subjects from the second file.

```{r subject_order}
knitr::kable(bd_out$samples, caption = "Subjects in merged file")
```

# SNP merge

A SNP merge combines files that cover different SNPs but the same (or
overlapping) set of subjects. The output contains all SNPs and the subjects
common to every input file.

The example splits the 10-SNP file into two 5-SNP files using `subsetbd`, then
merges them back together.

```{r snp_merge, message = FALSE, warning = FALSE}
bdose_snp_a   <- file.path(tempdir(), "snp_a.bdose")
bdose_snp_b   <- file.path(tempdir(), "snp_b.bdose")
bdose_snp_out <- file.path(tempdir(), "merged_snps.bdose")

locs <- bd_full$snps$location

subsetbd(bdfiles    = bdose_full,
         bdose_file = bdose_snp_a,
         locations  = locs[1:5])

subsetbd(bdfiles    = bdose_full,
         bdose_file = bdose_snp_b,
         locations  = locs[6:10])

mergebd(bdose_files = c(bdose_snp_a, bdose_snp_b),
        bdose_file  = bdose_snp_out)

bd_snp_a   <- getbdinfo(bdose_snp_a)
bd_snp_b   <- getbdinfo(bdose_snp_b)
bd_snp_out <- getbdinfo(bdose_snp_out)

cat("File A SNPs:    ", nrow(bd_snp_a$snps),   "\n")
cat("File B SNPs:    ", nrow(bd_snp_b$snps),   "\n")
cat("Merged SNPs:    ", nrow(bd_snp_out$snps),  "\n")
cat("Merged subjects:", nrow(bd_snp_out$samples), "\n")
```

## Verifying SNP order

SNPs appear in input-file order: all SNPs from the first file followed by all
SNPs from the second file.

```{r snp_order}
knitr::kable(bd_snp_out$snps, caption = "SNPs in merged file")
```

```{r cleanup, include = FALSE}
unlink(c(bdose_full,    paste0(bdose_full,    ".bdi"),
         bdose_a,       paste0(bdose_a,       ".bdi"),
         bdose_b,       paste0(bdose_b,       ".bdi"),
         bdose_out,     paste0(bdose_out,     ".bdi"),
         bdose_snp_a,   paste0(bdose_snp_a,   ".bdi"),
         bdose_snp_b,   paste0(bdose_snp_b,   ".bdi"),
         bdose_snp_out, paste0(bdose_snp_out, ".bdi")))
```
